---
title: "DEG_GFP_Tessera"
output: html_document
date: "2026-05-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(TESSERA)
library(BiocFileCache)
library(SummarizedExperiment)
library(SingleCellExperiment)
library(SpatialExperiment)
library(dplyr)
library(reshape2)
library(Matrix)
library(MatrixGenerics)
library(MASS)
library(fdrtool)
library(ggplot2)
```

#Data preparation
```{r}
bin16_list_ann <- readRDS("~/bin16_list_ann.rds")

#per capire meglio mi adeguo alla vignetta e tengo più condizioni assieme
treat_list <- bin16_list_ann[names(bin16_list_ann) %in%
                                   c("c26murf1_b2", "c26murf1_b3", "c26murf1_b5", "c26murf1_b9",
                                     "c26foxO_b1","c26foxO_b4","c26foxO_b6","c26foxO_b9",
                                     "c26SMAD23_b2","c26SMAD23_b4","c26SMAD23_b5","c26SMAD23_b9")]
treat_list$c26murf1_b2$to_discard <- NULL
treat_list$c26SMAD23_b2$to_discard <- NULL
treat_list$c26murf1_b2$in_treatment <- ifelse(treat_list$c26murf1_b2$in_treatment == TRUE, 1, 0)
treat_list$c26SMAD23_b2$in_treatment <- ifelse(treat_list$c26SMAD23_b2$in_treatment == TRUE, 1, 0)

myonuclei_types <- c("Myonuclei_IIb", "Myonuclei_IIx",
                     "Myonuclei_IIx_IIa", "Myonuclei_IIx_IIb",
                     "Myonuclei_MTJ", "Myonuclei_NMJ", "Myonuclei_Trim63")

for (nm in names(treat_list)) {
    spe <- treat_list[[nm]]
    colData(spe)$sample_id <- rep(nm, ncol(spe))
    treat_list[[nm]] <- spe
}

combined_spe <- do.call(cbind, treat_list)
combined_spe$in_treatment <- ifelse(combined_spe$in_treatment == 1, "gfp+", "untreated")
spe <- combined_spe[, combined_spe$new_cell_type %in% myonuclei_types]

# Add the spatial coordinates into the metadata (convenience)
colData(spe) <- cbind(colData(spe),spatialCoords(spe))

#fixing non-unique colnames
colnames(spe) <- paste0(colnames(spe), "_", spe$sample_id)
rownames(colData(spe)) <- colnames(spe) 

```

```{r}
rv <- rowVars(counts(spe))
names(rv) <- rownames(spe)
# Order the entire object by raw count variance (highest to lowest)
spe <- spe[order(rv, decreasing = TRUE), ]
```

```{r}
# Define covariate columns in metadata
celltype_col <- "new_cell_type"
sample_col <- "sample_id"
gfp_col <- "in_treatment" #aggiungo, da capire poi dove metterla
group_col <- "tissue_type"
coord_cols <- c("pxl_row_in_fullres", "pxl_col_in_fullres")
covariates_keep <- c(group_col, gfp_col) #NB: uso gfp al posto di celltype

# Define reference levels (baselines) for each factor (celltype, sample, and condition)
# ref_celltype <- ""
# ref_sample <- ""
# ref_group <- ""

# Re-level factors to ensure the model uses the correct baseline
# colData(spe)[, sample_col] <- relevel(factor(colData(spe)[, sample_col]), ref = ref_sample)
# colData(spe)[, celltype_col] <- relevel(factor(colData(spe)[, celltype_col]), ref = ref_celltype)
# colData(spe)[, group_col] <- relevel(factor(colData(spe)[, group_col]), ref = ref_group)

```

Multi-sample studies often require controlling for “sample effects” while samples are nested within broader experimental conditions To account for these effects without introducing multi-collinearity, we assign a nested ID (1, 2, 3…) to each sample within its respective condition.
```{r}
spe$tissue_type <- droplevels(spe$tissue_type)

# Extract unique sample IDs for each experimental condition
murf_ids <- unique(colData(spe)[, sample_col][colData(spe)[, group_col]  == "c26murf1"])
smad_ids <- unique(colData(spe)[, sample_col][colData(spe)[, group_col]  == "c26SMAD23"])
foxO_ids <- unique(colData(spe)[, sample_col][colData(spe)[, group_col]  == "c26foxO"])

# Initialize a new ID vector
new_id <- rep(0, nrow(colData(spe)))

# Assign a numeric index to samples within each condition
for (idx in 1:length(murf_ids)) {
  new_id[colData(spe)[, sample_col] == murf_ids[idx]] <- idx
}
for (idx in 1:length(smad_ids)) {
  new_id[colData(spe)[, sample_col] == smad_ids[idx]] <- idx
}
for (idx in 1:length(foxO_ids)) {
  new_id[colData(spe)[, sample_col] == foxO_ids[idx]] <- idx
}

# Store the nested ID back into the spatial experiment object
colData(spe)$nested_id <- factor(new_id)
new_nested_sample_col <- "nested_id"
#table(spe$nested_id,spe$tissue_type)
   
```

```{r}
# Build the formula string: ~ 0 + Group:celltype + Group:nested_id
design_formula <- as.formula(paste0(" ~ 0 + ", 
                                    group_col, ":", gfp_col, " + ", 
                                    group_col, ":nested_id"))
print(design_formula)
# Create the design matrix
X_mat <- model.matrix(design_formula, colData(spe))
head(X_mat)

X_mat <- X_mat[rownames(colData(spe)), ]
identical(rownames(X_mat),rownames(colData(spe)))

```
# Ensure the design matrix is full rank
```{r}
if (rankMatrix(X_mat)[1] != ncol(X_mat)) {
  stop("The design matrix is not full rank. Check for linear dependencies.")
}
dim(X_mat)
head(colnames(X_mat))
```

#Lattice model for spatial correlation matrix
For structured grid-based datasets, such as next-generation sequencing datasets from the Visium platform, you can simply provide the expected_num_neighbors argument (typically 6 for Visium due to its hexagonal grid). TESSERA then automatically determines the distance threshold required to satisfy the target neighbor count, defining any cells within this distance as adjacent.
Leroux is our preferred default; it decomposes variance into spatially correlated and uncorrelated components and offers the fastest pre-computation.

# Prepare data for the TESSERA package
```{r}
?TESSERA::prep_data
rownames(X_mat) <- rownames(colData(spe))
identical(rownames(X_mat),rownames(colData(spe)))
identical(rownames(colData(spe)), colnames(counts(spe)))

cm <- as.matrix(counts(spe)) 
colnames(cm) <- colnames(spe)
metadata = colData(spe)
coords = spatialCoords(spe) #PROBLEMA NOMI
rownames(coords) = colnames(spe)
  
TESSERA_data <- TESSERA::prep_data(
  x = cm,
  meta_data = metadata,
  sample_col = sample_col,
  coord_data = coords,
  design_mat = X_mat,
  model_type = "Leroux",
  expected_num_neighbors = 6
)

```

#Run TESSERA
NB: METTERE IN PARALLELO -> VEDI VIGNETTA!!!

By inspecting the fitted spatial covariance parameters, we can quantify the strength of the spatial dependence for the lattice models via the identifiable parameters gamma_hat or tau2_hat. 
The MSE results, such as MSE_counts_sample relative to Mean_counts2_sample, facilitate the evaluation of fitted count accuracy on a per-sample basis.
Finally, the Moran’s I metrics, specifically Moran_residuals, act as a critical diagnostic to ensure that the spatial structure has been adequately captured, indicating that the model has successfully accounted for the spatial dependency in the data.
```{r}
# Which gene to run
gene_idx <- which(rownames(spe) == "Trim63")

TESSERA_out <- TESSERA::TESSERA_lattice(
  TESSERAData_obj = TESSERA_data,
  gene_name = rownames(spe)[gene_idx],
  model_type = "Leroux",
)

print(TESSERA_out$time)
head(TESSERA_out$performanceSummary, 2)
head(TESSERA_out$beta_hat)
```

After fitting models to hundreds or thousands of genes, we aim to identify which among them are differentially expressed. 

