---
title: "Neighborhood analysis nuclei"
output: html_document
date: "2026-01-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(SpatialExperiment)
library(SpatialFeatureExperiment)
library(purrr)
library(imcRtools)
library(dplyr)
library(tidyr)
library(ggplot2)
library(patchwork)
library(RColorBrewer)
library(BiocParallel)
library(SpatialExperiment)
library(cluster)
 library(ggspavis)
```

#Reading nuclei data
```{r}
sfe_list <- readRDS("~/sfe_list_annotated")
filtered_sfe <- sfe_list |> 
  map(function(sfe){
    sfe <- sfe[, !sfe$to_discard]
    sfe
})
# labels name colData: "labels_singler"
```

#Neighborhood analysis with imcRtools as described in https://bioconductor.org/books/3.22/OSTA/pages/img-neighborhood-analysis.html
```{r}
#spe <- sfe_list$blocco1_sham
spe <- filtered_sfe$blocco1_sham

# construct kNN-graph based on Euclidean distances
sqe <- buildSpatialGraph(spe, 
                         coords=spatialCoordsNames(spe),
                         img_id="sample_id",
                         type="knn", k=10) #anche qua potrei cambiare k e vedere se migliora
# compute cluster frequencies among each cell's kNNs
sqe <- aggregateNeighbors(sqe, 
                          colPairName="knn_interaction_graph", 
                          aggregate_by="metadata", count_by="labels_singler")
# view composition of 1st cell's kNNs
unlist(sqe$aggregatedNeighbors[1, ]) 
```

```{r}
# cluster cells by neighborhood compositions
#ctx <- kmeans(sqe$aggregatedNeighbors, centers=5) #eh... come lo decido? biologi? silhouette?

set.seed(2753)
ctx <- kmeans(sqe$aggregatedNeighbors, centers=10) 
table(sqe$ctx <- factor(ctx$cluster))
# dist_mat <- dist(sqe$aggregatedNeighbors)
# sil <- silhouette(ctx$cluster, dist = dist_mat)
# mean(sil[, "sil_width"])

```

#Let’s quickly view the subpopulation composition of each spatial context:
```{r}
df <- data.frame(spatialCoords(sqe), colData(sqe))
round(100*with(df, prop.table(table(labels_singler, ctx), 2)), 2)
```

#visualization
```{r}
plotCoords(sqe, annotate = "ctx")
```

#provo scider
```{r}
#BiocManager::install("scider")
library(scider)
library(SpatialExperiment)
```

#density calculation
```{r}
sfe_list <- readRDS("~/sfe_list_annotated")
filtered_sfe <- sfe_list |> 
  map(function(sfe){
    sfe <- sfe[, !sfe$to_discard]
    sfe
})
spe <- filtered_sfe$blocco1_sham
#gridDensity si aspetta una colonna cell_type
spe$cell_type <- spe$labels_singler
spe <- gridDensity(spe)
names(metadata(spe))
metadata(spe)$grid_density #ha calcolato la densità per tutti i tipi cellulari
```

```{r}
plotDensity(spe,coi="Myonuclei_IIx_IIb")
```

#After obtaining grid-based density for each COI, we can then detect regions-of-interest (ROIs) based on density or select by user.
spe <- findROI(spe, coi = "Fibroblasts")
metadata(spe)$fibroblasts_roi


#BANKSY -------------------------------------------------------------------------------------------------------
```{r}
library(Banksy)
library(SummarizedExperiment)
library(SpatialExperiment)
library(scuttle)
library(scater)
library(cowplot)
library(ggplot2)
```

```{r}

# sfe_list <- readRDS("~/sfe_list_annotated")
# filtered_sfe <- sfe_list |> 
#   map(function(sfe){
#     sfe <- sfe[, !sfe$to_discard]
#     sfe
# })
# spe <- filtered_sfe$blocco1_sham
#spe <- filtered_sfe$blocco2_c26

#SUI 16
load("~/subset_list.RData")
spe <- subset_list$sham_b1

# QC based on total counts
qcstats <- perCellQCMetrics(spe)
thres <- quantile(qcstats$total, c(0.05, 0.98))
keep <- (qcstats$total > thres[1]) & (qcstats$total < thres[2])
spe <- spe[, keep] #circa 2000 nuclei eliminati

# Normalization to mean library size
spe <- computeLibraryFactors(spe)
aname <- "normcounts"
assay(spe, aname) <- normalizeCounts(spe, log = FALSE)
```

#DOMAINS
```{r}
#parametri suggeriti per domain segmentation con visiumHD
lambda <- 0.8
k_geom <- 30
npcs <- 50

set.seed(1000)
spe <- computeBanksy(spe, assay_name = aname, k_geom = k_geom)
spe <- runBanksyPCA(spe, lambda = lambda, npcs = npcs)
spe <- clusterBanksy(spe, lambda = lambda, npcs = npcs, resolution = 0.5) #abbasso la resolution da 0.8

spe$domains <- spe$clust_M0_lam0.8_k50_res0.5 #cambiare anche qua
table(spe$domains)

```

#visualizzazione
```{r}
col <- hcl.colors(n = nlevels(spe$domains), palette = "Dark 3")
plotCoords(spe, annotate = "domains")

```

#Faccio Banksy direttamente sugli 8bin visto che non parte da un'annotazione cmq.
```{r}
load("~/spe_list_8bin_ann2.RData")
spe <- spe_list$sham_b1
keep <- !is.na(spe_list$sham_b1$cell_type)
spe <- spe[,keep]

# QC based on total counts
qcstats <- perCellQCMetrics(spe)
thres <- quantile(qcstats$total, c(0.05, 0.98))
keep <- (qcstats$total > thres[1]) & (qcstats$total < thres[2])
spe <- spe[, keep] #circa 2000 nuclei eliminati

# Normalization to mean library size
spe <- computeLibraryFactors(spe)
aname <- "normcounts"
assay(spe, aname) <- normalizeCounts(spe, log = TRUE)

#prendo solo i geni + variabili se no non ce la fa
library(scran)
dec <- modelGeneVar(spe, assay.type = "normcounts")
hvg <- getTopHVGs(dec, n = 2000)
spe <- spe[hvg, ]

```

```{r}
#parametri suggeriti per domain segmentation con visiumHD
lambda <- 0.8
k_geom <- 30
npcs <- 50

set.seed(1000)
spe <- computeBanksy(spe, assay_name = aname, k_geom = k_geom)
spe <- runBanksyPCA(spe, lambda = lambda, npcs = npcs)
spe <- clusterBanksy(spe, lambda = lambda, npcs = npcs, resolution = 0.2)

spe$domains <- spe$clust_M0_lam0.8_k50_res0.2
table(spe$domains)

```

```{r}
plotCoords(spe, annotate = "domains")

```
