---
title: "alternative_deconvolutions"
output: html_document
date: "2025-12-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(STdeconvolve)
library(SpatialExperiment)
library(SMART)
library(readxl)
library(ggspavis)
```

#ScType
ScType workflow consists of the following steps:
Input data quality control (as implemented in Seurat pipeline, latest version) to visually explore the data, including:
Cell- based quality control (e.g. removing cells with unexpectedly high counts)
Gene- based quality control (i.e. subsetting highly-variable genes)
Data normalization and scaling (The default normalization is done using SCTransform, but other methods, such as log-normalization are available)
Data clustering based on highly variable genes (default is Louvain clustering based on a shared nearest neighbor graph, but many other options are available)
Differential expression analysis and cell marker identification.
Automatic cell type identification and quantification
Visualization of results (multiple options are available)
```{r}
#non è scaricabile il pacchetto
# lapply(c("dplyr", "Seurat", "HGNChelper", "openxlsx"), library, character.only = TRUE)
# source("https://raw.githubusercontent. com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R")
# source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_. R")
# source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R")
# source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")

load("~/spe_list_8bin.RData")
spe <- spe_list$blocco1_sham

#converti a Seurat
library(Seurat)
counts_mat <- as.matrix(assay(spe, "X"))
cell_metadata <- as.data.frame(colData(spe))
seu <- CreateSeuratObject(
  counts = counts_mat,
  meta.data = cell_metadata
)
save(seu,file="sham_b1.RData")

#HACTAR

#controllo qualità e normalizzazione
# seu<- subset(seu, subset = nFeature_RNA > 200 & nFeature_RNA < 2500) 
# plot(density(Matrix::colSums(seu@assays$RNA@layers$counts > 0)),xlim=c(0,100))
#intanto non tolgo nulla, al massimo cmq meglio togliere con gli UMI

seu <- NormalizeData(seu, normalization.method = "LogNormalize", scale.factor = 10000)
seu <- FindVariableFeatures(seu, selection.method = "vst", nfeatures = 5000) #alzo un po'

# scale and run PCA
seu <- ScaleData(seu, features = rownames(seu))
seu <- RunPCA(seu, features = VariableFeatures(object = seu))

# Check number of PC components (we selected 10 PCs for downstream analysis, based on Elbow plot)
# ElbowPlot(seu)

# cluster and visualize
seu <- FindNeighbors(seu, dims = 1:10)
seu <- FindClusters(seu, resolution = 0.8)
seu <- RunUMAP(seu, dims = 1:10)

load("~/seu.RData")
DimPlot(seu, reduction = "umap")
#come previsto, fa schifo

```

#STdeconvolve
```{r}
load("~/spe_list_8bin.RData")
spe <- spe_list$blocco1_sham
## remove pixels with too few genes
?cleanCounts
X_mat <- as.matrix(assay(spe, "X"))
counts <- cleanCounts(counts = X_mat, 
                      min.lib.size = 50, #proviamo intanto così
                      min.reads = 1,
                      min.detected = 1,
                      verbose = TRUE)

?restrictCorpus
#finds hvg
corpus <- restrictCorpus(counts,
                         removeAbove = 1.0,
                         removeBelow = 0.05,
                         alpha = 0.05,
                         plot = TRUE,
                         verbose = TRUE)
#ne rimangono solo 70 così, bah

# STdeconvolve then applies latent Dirichlet allocation (LDA), a generative statistical model
# commonly used in natural language processing, to discover K latent cell-types.
# STdeconvolve fits a range of LDA models to inform the choice of an optimal K.
## Note: the input corpus needs to be an integer count matrix of pixels x genes
ldas <- fitLDA(t(as.matrix(corpus)), Ks = seq(2, 9, by = 1),
               perc.rare.thresh = 0.05,
               plot=TRUE,
               verbose=TRUE)
## select model with minimum perplexity
optLDA <- optimalModel(models = ldas, opt = "min")
#k=2
#fa schifo


```

#SMART
Provo a fare con i marker aggiornati?
```{r}
devtools::install_github("yyolanda/SMART")

load("~/spe_list_8bin.RData")
spe <- spe_list$blocco1_sham
counts = t(counts(spe))
markers_list <- readRDS("~/marker_genes_list_singleR_annot.rds")

#Controllo qualità (stessa soglia di RCTD)
counts <- counts[rowSums(counts) > 15,]
counts <- counts[ ,colSums(counts) > 0]
#per dopo
spe <- spe[,colSums(counts(spe))>15]
spe <- spe[rowSums(counts(spe))>0,]

#controlla quali markers non sono espressi
markers_list <- lapply(markers_list, function(v) v[v %in% colnames(counts)])

base_res <- SMART_base(stData = counts,
                       markerGs = markers_list, 
                       noMarkerCts = 0, #num. tipi cell. senza markers
                       #outDir='SMART_results',
                       seed=8001)

cell_type <- apply(base_res$ct_proportions, 1, function(x) colnames(base_res$ct_proportions)[which.max(x)])
table(cell_type)
base_res$cell_type <- cell_type
spe$cell_type <- base_res$cell_type

your_color_vector <- c(
    "Endothelial" = "#0072B2",       # blu vivo
    "FAPs" = "#009E73",              # verde vivo
    "Immune_Cells" = "#D55E00",      # arancione-rosso
    "MuSC" = "#E69F00",              # arancione
    "Myonuclei_IIb" = "#CC79A7",     # magenta
    "Myonuclei_IIx" = "#56B4E9",     # azzurro chiaro
    "Myonuclei_IIx_IIa" = "#F0E442", # giallo
    "Myonuclei_IIx_IIb" = "#9B59B6", # viola chiaro
    "Myonuclei_MTJ" = "#0099B4",     # turchese
    "Myonuclei_NMJ" = "#DDAA33",     # senape
    "Myonuclei_Trim63" = "#CC3311",  # rosso scuro
    "Nervous_System" = "#44AA99",    # verde acqua
    "Pericyte" = "#AA4499",          # viola/rosa
    "Smooth_Muscular" = "#332288",   # blu-viola scuro
    "Tenocyte" = "#A3E635"           # verde lime
)

stat1 <- plotCoords(spe, annotate = "cell_type", point_size = 0.01) +
    scale_color_manual(values = your_color_vector) +
    theme(
      legend.key.width = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
      )


```

#SpiceMix (python)
```{r}
library(reticulate)

# Create conda env with Python 3.10
reticulate::conda_create("spicemix", packages = "python=3.10")

# Install PyTorch: prefer conda/pytorch channels for correct wheel; example (CPU) via pip index:
reticulate::conda_install("spicemix", packages = c("pip"), channel = "defaults")

# Use pip (reticulate wraps pip) to install torch and spicemix into the conda env
reticulate::py_install(
  packages = c(
    "--index-url", "https://download.pytorch.org/whl/cpu", "torch",
    "spicemix", "anndata", "numpy", "scipy"
  ),
  envname = "spicemix",
  pip = TRUE
)

```

#CARD
https://bioconductor.posit.co/packages/release/bioc/vignettes/CARDspa/inst/doc/Example_Analysis.html
We can use CARD_refFree to do reference-free deconvolution. This function frist creat a CARDfree object and then do deconvolution. Briefly, the essential inputs are the same as the function CARD_deconvolution, except that this function does not require the single cell count and meta information matrix. Instead, it requires a markerList.
```{r}
library(CARDspa)
library(RcppML)
library(NMF)
library(RcppArmadillo)
library(SingleCellExperiment)
library(SpatialExperiment)
library(ggplot2)

load("~/markers_list_mixed.RData") #lista aggiornata
load("~/spe_list_8bin.RData")
spe <- spe_list$blocco1_sham
rownames(spatialCoords(spe)) <- colnames(spe)
coords_spe <- as.data.frame(spatialCoords(spe))

?CARD_refFree
set.seed(seed = 20200107)
CARDfree_obj <- CARD_refFree(
    markerlist = markers_list,
    spatial_count = counts(spe),
    spatial_location = coords_spe,
    mincountgene = 100,
    mincountspot = 5
)
## STOP! The average number of unique marker genes for each cell
# type is less than 20 ...

```

#SpatialDWLS
Fa parte di GIOTTO
PROBLEMA PACCHETTI PYTHON
```{r}
library(Giotto)
load("~/spe_list_8bin.RData")
sham_b1 <- spe_list$blocco1_sham

my_giotto_object = createGiottoObject(raw_exprs = counts(sham_b1),
                                      spatial_locs = as.data.frame(spatialCoords(sham_b1)))

#filtering
?filterGiotto
my_giotto_object <- filterGiotto(gobject = my_giotto_object, 
                                 expression_threshold = 1, #prova ad abbassare
                                 gene_det_in_min_cells = 3, 
                                 min_det_genes_per_cell = 10)

#normalization
?normalizeGiotto
my_giotto_object <- normalizeGiotto(gobject = my_giotto_object, scalefactor = 6000, verbose = T)

# dimension reduction
my_giotto_object <- calculateHVG(gobject = my_giotto_object)
my_giotto_object <- runPCA(gobject = my_giotto_object)
my_giotto_object <- runUMAP(my_giotto_object, dimensions_to_use = 1:5)

#leiden
my_giotto_object <- createNearestNetwork(
  gobject = my_giotto_object,
  dimensions_to_use = 1:10,   # or however many PCA dims you want
  k = 15,
  name = "NN_network"
)
my_giotto_object = doLeidenCluster(my_giotto_object, name = 'leiden_clus', nn_network_to_use = "NN_network")

?runDWLSDeconv
runDWLSDeconv(
  gobject,
  expression_values = c("normalized"),
  logbase = 2,
  cluster_column = "leiden_clus",
  sign_matrix,
  n_cell = 50,
  cutoff = 2,
  name = NULL,
  return_gobject = TRUE
)
```

#SPOTlight
```{r}
BiocManager::install("SPOTlight")
library(SPOTlight)
library(Matrix)
library(data.table)
library(Seurat)
library(SeuratData)
library(dplyr)
library(gt)
library(SPOTlight)
library(igraph)
library(RColorBrewer)


```

#SingleR



