---
title: "16um bins annotation with Rctd"
author: "Emma Menna"
output: html_document
date: "2025-07-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Libraries
```{r, include=FALSE}
library(arrow)
library(Banksy)
library(BiocParallel)
library(DropletUtils)
library(dplyr)
library(edgeR)
library(ggplot2)
library(graphics)
library(igraph)
library(jsonlite)
library(Matrix)
library(OSTA.data)
library(patchwork)
library(pheatmap)
library(S4Vectors)
library(scDblFinder)
library(scater)
library(scran)
library(scuttle)
library(SingleCellExperiment)
library(SpatialExperiment)
library(SummarizedExperiment)
library(tidyr)
library(VisiumIO)
library(spacexr)
library(SpotSweeper)
library(ggspavis)
library(writexl)
library(lme4)
```

#Objects
```{r}
sce_ref <- readRDS("~/multiome_sce.rds") #reference dataset
subset_list <- readRDS("~/bin16_list_raw.rds")

color_vector <- c(
    "Endothelial" = "#0072B2",       # blu vivo
    "FAPs" = "#009E73",              # verde vivo
    "Immune_Cells" = "#D55E00",      # arancione-rosso
    "MuSC" = "#E69F00",              # arancione
    "Myonuclei_IIb" = "#CC79A7",     # magenta
    "Myonuclei_IIx" = "#56B4E9",     # azzurro chiaro
    "Myonuclei_IIx_IIa" = "#F0E442", # giallo
    "Myonuclei_IIx_IIb" = "#9B59B6", # viola chiaro
    "Myonuclei_MTJ" = "#0099B4",     # turchese
    "Myonuclei_NMJ" = "#DDAA33",     # senape
    "Myonuclei_Trim63" = "#CC3311",  # rosso scuro
    "Nervous_System" = "#44AA99",    # verde acqua
    "Pericyte" = "#AA4499",          # viola/rosa
    "Smooth_Muscular" = "#332288",   # blu-viola scuro
    "Tenocyte" = "#A3E635"           # verde lime
)

score_cols <- c("Endothelial", "FAPs", "Immune_Cells", "MuSC", 
                "Myonuclei_IIb", "Myonuclei_IIx", "Myonuclei_IIx_IIa", 
                "Myonuclei_IIx_IIb", "Myonuclei_MTJ", "Myonuclei_NMJ", 
                "Myonuclei_Trim63", "Nervous_System", "Pericyte", 
                "Smooth_Muscular", "Tenocyte")

```

#Deconvolution function
For each spot the probability for each cell type as well as the label of the principal cell type are saved.
Three assays (one in full mode):
weights: Cell type proportions restricted according to the specified mode
weights_unconfident: Cell type proportions restricted according to the specified mode, including unconfident predictions (not available in full mode)
weights_full: Unrestricted cell type proportions (not available in full mode, use weights instead)
Assays have cell types as rows and pixels as columns, with values representing the proportion (0 to 1) of each cell type in each pixel. Assay columns sum to 1 (except for rejected pixels, which sum to 0).
```{r}
run_my_rctd <- function(spe, sce_ref, out_prefix) {
  set.seed(9237)
  rownames(spe) <- rowData(spe)$Symbol
  rownames(spe) <- make.unique(rownames(spe))
  
  #data preparation
  rctd_data <- createRctd(
    spatial_experiment = spe,
    reference_experiment = sce_ref,
    cell_type_col = "subclusters"
  )
  
# For multi mode:
# cell_type_list: List of cell types per pixel
# conf_list: List of whether cell type predictions are confident
# Additional metrics like min_score (likelihood)
  
  #deconvolution
  results <- runRctd(rctd_data, rctd_mode = "multi")

  #weights extraction
  ws <- assay(results, "weights")
  ws <- data.frame(t(as.matrix(ws)))
  colData(spe)[names(ws)] <- ws[colnames(spe), ] 
  
  #Cluster labelling
  table(rowSums(ws == 1))
  ids <- names(ws)[apply(ws, 1, which.max)]
  ids <- gsub("\\.([A-z])", " \\1", ids)
  idx <- match(colnames(spe), rownames(ws))
  spe$cell_type <- factor(ids[idx]) 
  print(table(spe$cell_type))
  
  assign("subset_list", replace(subset_list, out_prefix, spe), envir = .GlobalEnv)
  invisible(spe)
}
```

#Function to fix cell types
Rctd in "multi" mode sometimes assigns probabilities all equal to 0
then the function above wrongly assignes the first cell type in the list.
```{r}
fix_spe_celltypes <- function(spe, score_columns, cell_type_col = "cell_type") {

  cd <- as.data.frame(colData(spe))
  score_mat <- as.matrix(cd[, score_columns])
  max_score <- apply(score_mat, 1, max)
  best_type <- score_columns[max.col(score_mat, ties.method = "first")]
  orig_col_idx <- match(as.character(cd[[cell_type_col]]), score_columns)
  orig_score <- score_mat[cbind(1:nrow(score_mat), orig_col_idx)]
  
  new_cell_type <- as.character(cd[[cell_type_col]])
  
  # If original score is 0, but another type is > 0 -> Reassign to best_type
  reassign_mask <- which(orig_score == 0 & max_score > 0)
  new_cell_type[reassign_mask] <- best_type[reassign_mask]
  
  # If EVERY score is 0 -> Assign NA
  na_mask <- which(max_score == 0)
  new_cell_type[na_mask] <- NA

  spe$new_cell_type <- new_cell_type
  return(spe)
}

```

#Calling deconvolution function for each sample
```{r}
for (name in names(subset_list)) {
    run_my_rctd(subset_list[[name]], sce_ref, name)
}
```

#Fixing rctd results
```{r}
#1. removing spots with null probabilities
subset_list <- map(subset_list, \(spe) {spe <- fix_spe_celltypes(spe, score_cols)})
map(subset_list, \(x){table(x$new_cell_type)})

#2. removing cells with NA cell type
subset_list <- lapply(subset_list, function(spe) {
   keep <- !is.na(spe$new_cell_type) & colSums(counts(spe)) > 0
   spe[,keep]
})

#3. manual correction on MuSC annotation: defined as Pax7 expressed or not
subset_list <- lapply(subset_list, function(spe) {
  print(table(spe$new_cell_type == "MuSC"))
  pax7_pos <- counts(spe)["Pax7", ] > 0
  spe$new_cell_type[pax7_pos] <- "MuSC"
  print(table(spe$new_cell_type == "MuSC"))
  return(spe)
})

#saving results
saveRDS(subset_list, file = "bin16_list_ann.rds")
```

#Projections on tissue for all samples
For each sample 2 plots are created. In the first, for all cell types spots are colored according to the probabiluty (weigth) for that cell type. In the second one, each spot is colored according to its clustering labels (the cell type with highest weigth).
```{r}

for (spe_name in names(subset_list)) {
  spe <- subset_list[[spe_name]]
  ws <- colData(spe)[,seq(8,22,1)]
  p <- lapply(names(ws), function(nm) {
  plotCoords(spe, annotate = nm, point_size = 0.1) +
    scale_color_gradientn(
      colors = rev(hcl.colors(9, "Rocket")),
      limits = c(0, 1)  # fissa scala da 0 a 1
    ) +
    ggtitle(nm)
  })

  final_plot <- wrap_plots(p, nrow = 3, ncol = 5, guides = "collect") &
  theme(
    legend.key.width = unit(0.5, "lines"),
    legend.key.height = unit(1, "lines"),
    strip.text = element_text(size = 5)
  ) &
  scale_color_gradientn(colors = rev(hcl.colors(9, "Rocket")), limits = c(0, 1))

  ggsave(paste0(spe_name, "_types.png"), final_plot, width = 10, height = 8, dpi = 300)

stat1 <- plotCoords(spe, annotate = "new_cell_type", point_size = 0.5) +
    scale_color_manual(values = your_color_vector) +
    theme(
      legend.key.width = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    )

ggsave(paste0(out_prefix, "_dec.png"), stat1, width = 10, height = 8, dpi = 300)
}
  
```

#Barplot to show cell type proportions for each sample
```{r}
prop_df <- lapply(names(subset_list), function(name) {
  spe <- subset_list[[name]]
  tab <- table(spe$cell_type)
  prop <- prop.table(tab)
  data.frame(
    sample = name,
    cell_type = names(prop),
    proportion = as.numeric(prop)
  )
}) |> bind_rows()

ggplot(prop_df, aes(x = sample, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = your_color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per campione",
    x = "Campione (subset_list)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("cell_type_proportions.png", width = 10, height = 6, dpi = 300)

```

#Barplot to show cell types proportions for each condition
From this point on block 7 samples won't be considered anymore.
```{r}
list_wb7 <- subset_list[!names(subset_list) %in% c("c26_b7","c26STAT3_b7","sham_b7")]
prop_df_condition <- lapply(list_wb7, function(spe) {
  data.frame(
    tissue_type = unique(spe$tissue_type),
    cell_type = spe$cell_type
  )
}) |> bind_rows() |> filter(!is.na(cell_type))

prop_df_condition <- prop_df_condition %>%
  dplyr::count(tissue_type, cell_type) %>%
  dplyr::group_by(tissue_type) %>%
  dplyr::mutate(proportion = n / sum(n)) %>%
  ungroup()

ggplot(prop_df_condition, aes(x = tissue_type, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = your_color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per condizione",
    x = "Condizione (tissue_type)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("cell_type_proportions_by_condition.png", width = 10, height = 6, dpi = 300)

```

#Same barplot for multiome annotations for comparison
```{r}
table(colData(sce_ref)$condition)

prop_df_condition <- 
  data.frame(
    tissue_type = colData(sce_ref)$condition,
    cell_type =  colData(sce_ref)$subclusters
  )

prop_df_condition <- prop_df_condition %>%
  count(tissue_type, cell_type) %>%
  group_by(tissue_type) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup()

ggplot(prop_df_condition, aes(x = tissue_type, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = your_color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per condizione",
    x = "Condizione (tissue_type)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("cell_type_multiome.png", width = 10, height = 6, dpi = 300)

```

#Table to see proportions 
```{r}
#by sample
celltype_table <- lapply(names(subset_list), function(sample_name) {
  spe <- subset_list[[sample_name]]
  df <- as.data.frame(colData(spe))
  df <- df %>%
    group_by(cell_type) %>%
    summarise(count = n()) %>%
    mutate(proportion = count / sum(count),
           sample = sample_name)
  df
}) %>%
  bind_rows()

celltype_matrix <- celltype_table %>%
  select(sample, cell_type, proportion) %>%
  pivot_wider(names_from = cell_type, values_from = proportion, values_fill = 0)

celltype_matrix
#write_xlsx(celltype_matrix, "celltype_proportions_per_sample.xlsx")

#by condition
ct_long <- lapply(list_wb7, function(spe) {
  data.frame(
    tissue_type = unique(spe$tissue_type), 
    cell_type   = spe$cell_type
  )
}) %>%
  bind_rows()

counts <- ct_long %>%
  count(tissue_type, cell_type, name = "n")

props <- counts %>%
  group_by(tissue_type) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

props_wide <- props %>%
  select(tissue_type, cell_type, prop) %>%
  pivot_wider(names_from = cell_type, values_from = prop, values_fill = 0)

write_xlsx(props_wide, path = "celltype_by_tissue_type.xlsx")

```

#boxplot of cell type proportions stratified by condition
```{r}
prop_per_sample <- purrr::imap_dfr(subset_list, function(spe, name) {
  cd <- as.data.frame(SummarizedExperiment::colData(spe))
  cd |>
    filter(!is.na(cell_type)) |>
    dplyr::count(cell_type, name = "count") |>
    mutate(proportion  = count / sum(count),
           sample      = name,
           tissue_type = paste(unique(as.character(cd$tissue_type)), collapse = ", ")) |>
    dplyr::rename(cell_type = cell_type)
})
ggplot(prop_per_sample, aes(x = tissue_type, y = proportion, fill = tissue_type)) +
  geom_boxplot(outlier.shape = 21) +
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1B9E77", "c26" = "#D95F02")) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribuzione proporzioni cell_type per tissue_type",
       x = "Tissue type",
       y = "Proporzione")
ggsave("boxplot_cell_types.png", width = 10, height = 6, dpi = 300)

```

 