---
title: "Esplorazione Bin"
author: "Valerio Reffo"
date: "2026-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library, message=FALSE, warning=FALSE, include=FALSE}
library(purrr)
library(ggplot2)
library(ggspavis)
library(jsonlite)
library(SpatialExperiment)
library(data.table)
library(ggnewscale)
library(SingleCellExperiment)
library(SingleR)
library(BiocParallel)
library(dplyr)
library(tidyr)
library(forcats) # For ordering plots

```

```{r}
#load("~/subset_list.RData")
#spe <-subset_list$sham_b1

#marker_genes_list_singleR_annot <- readRDS("~/marker_genes_list_singleR_annot.rds")

color_vector <- c(
    "Endothelial" = "#0072B2",       # blu vivo
    "FAPs" = "#009E73",              # verde vivo
    "Immune_Cells" = "#D55E00",      # arancione-rosso
    "MuSC" = "#E69F00",              # arancione
    "Myonuclei_IIb" = "#CC79A7",     # magenta
    "Myonuclei_IIx" = "#56B4E9",     # azzurro chiaro
    "Myonuclei_IIx_IIa" = "#F0E442", # giallo
    "Myonuclei_IIx_IIb" = "#9B59B6", # viola chiaro
    "Myonuclei_MTJ" = "#0099B4",     # turchese
    "Myonuclei_NMJ" = "#DDAA33",     # senape
    "Myonuclei_Trim63" = "#CC3311",  # rosso scuro
    "Nervous_System" = "#44AA99",    # verde acqua
    "Pericyte" = "#AA4499",          # viola/rosa
    "Smooth_Muscular" = "#332288",   # blu-viola scuro
    "Tenocyte" = "#A3E635"           # verde lime
)

```

```{r}
spe
```

# Analysis of RCTD results
There are a lot of Endothelial, but why? what is the distribution of the probability for each bin to be a Endothetial? 

```{r}
cd_df <- colData(spe) |> as_tibble()

cd_df |> 
  filter(cell_type == "Endothelial") |> 
  summarise(mean = mean(Endothelial),
            sd = sd(Endothelial),
            min = min(Endothelial),
            max = max(Endothelial))

cd_df |> 
  filter(cell_type == "Endothelial", Endothelial == 0) |> 
  summarise(n = n())
cd_df |> 
  filter(cell_type == "Endothelial", Endothelial == 1) |> 
  summarise(n = n())
cd_df |> 
  filter(
    cell_type == "Endothelial", 
    between(Endothelial, 0.5, 0.99)
  ) |> 
  summarise(n = n())
cd_df |> 
  filter(
    cell_type == "Endothelial", 
    between(Endothelial, 0.0001, 0.5)
  ) |> 
  summarise(n = n())
#ce ne sono pochissime tra 0 e 0.5

```
#Stessa cosa per un altro tipo cellulare
```{r}

cd_df |> 
  filter(cell_type == "Myonuclei_IIb") |> 
  summarise(mean = mean(Myonuclei_IIb),
            sd = sd(Myonuclei_IIb),
            min = min(Myonuclei_IIb),
            max = max(Myonuclei_IIb))

cd_df |> 
  filter(cell_type == "Myonuclei_IIb", Myonuclei_IIb == 0) |> 
  summarise(n = n())
cd_df |> 
  filter(cell_type == "Myonuclei_IIb", Myonuclei_IIb == 1) |> 
  summarise(n = n())
cd_df |> 
  filter(
    cell_type == "Myonuclei_IIb", 
    between(Myonuclei_IIb, 0.5, 0.99)
  ) |> 
  summarise(n = n())
cd_df |> 
  filter(
    cell_type == "Myonuclei_IIb", 
    between(Myonuclei_IIb, 0.0001, 0.5)
  ) |> 
  summarise(n = n())

#STESSO DISCORSO

```

```{r}
# 1. Filter only the bins assigned as Endothelial
p_hist <- as.data.frame(cd_df) |> 
  filter(cell_type == "Endothelial") |>
  ggplot(aes(x = Endothelial)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "black", alpha = 0.8) +
  theme_minimal() +
  labs(
    title = "Distribution of Endothelial Scores",
    subtitle = "Only including bins classified as 'Endothelial'",
    x = "Endothelial Score (0-1)",
    y = "Number of Bins"
  )

print(p_hist)
```


Second highest scores among Endothelial defined cell types
```{r}
# Define all your cell type score columns
score_cols <- c("Endothelial", "FAPs", "Immune_Cells", "MuSC", 
                "Myonuclei_IIb", "Myonuclei_IIx", "Myonuclei_IIx_IIa", 
                "Myonuclei_IIx_IIb", "Myonuclei_MTJ", "Myonuclei_NMJ", 
                "Myonuclei_Trim63", "Nervous_System", "Pericyte", 
                "Smooth_Muscular", "Tenocyte")

# Extract the second highest score for each Endothelial bin
second_highest_df <- cd_df |> 
  filter(cell_type == "Endothelial", Endothelial != 0) |>
  select(location_id, all_of(score_cols)) |> 
  pivot_longer(cols = -location_id, names_to = "candidate_type", values_to = "score") |>
  # Group by each individual bin
  group_by(location_id) |>
  # Sort scores from highest to lowest inside each bin
  arrange(desc(score)) |>
  # Grab the 2nd row (which is the second highest score)
  slice(2) |> 
  ungroup()

# Plot: WHICH cell types are most frequently the "second choice"?
p_2nd_identity <- ggplot(second_highest_df, aes(x = fct_infreq(candidate_type))) +
  geom_bar(fill = "darkorange", color = "black", alpha = 0.8) +
  theme_minimal() +
  coord_flip() + # Flip for readable names
  labs(
    title = "What is sharing space with Endothelial cells?",
    subtitle = "Most common 'second-highest' cell types in Endothelial bins",
    x = "Cell Type",
    y = "Count of Bins"
  )

print(p_2nd_identity)

```

```{r}
plot_confidence <- cd_df |> 
  filter(cell_type == "Endothelial", Endothelial != 0) |>
  left_join(second_highest_df |> select(location_id, second_type = candidate_type, second_score = score), by = "location_id") |>
  mutate(Confidence_Delta = Endothelial - second_score) |> 
  ggplot(aes(x = Confidence_Delta)) +
  geom_histogram(bins = 30, fill = "purple", color = "black", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Assignment Confidence (Endothelial Score - 2nd Highest Score)",
       x = "Delta (Higher = Purer Bin, Lower = Mixed Bin)")
plot_confidence
```

```{r}
weird_endo <- cd_df |> 
  filter(cell_type == "Endothelial", Endothelial == 0)

# 2. Check the MAXIMUM score for all other cell types in these bins
cat("\n--- MAXIMUM SCORE IN OTHER CELL TYPES ---\n")
weird_endo |> 
  select(FAPs:Tenocyte) |> 
  summarise(across(everything(), max)) |> 
  print(width = Inf)

# 3. Check HOW MANY of these 5700 bins have a score > 0 for the other cell types
cat("\n--- NUMBER OF BINS WITH SCORE > 0 ---\n")
weird_endo |> 
  select(FAPs:Tenocyte) |> 
  summarise(across(everything(), ~ sum(. > 0))) |> 
  print(width = Inf)
```

```{r}
# 2. Reshape and calculate the summary
zero_score_summary <- cd_df |> 
  # Keep only the essential columns (location_id, assigned cell_type, and the score columns)
  select(location_id, assigned_type = cell_type, all_of(score_cols)) |> 
  
  # Pivot the score columns into two new columns: 'score_type' and 'score_value'
  pivot_longer(
    cols = all_of(score_cols), 
    names_to = "score_type", 
    values_to = "score_value"
  ) |> 
  
  # KEEP ONLY the rows where the assigned type matches the score column name
  filter(assigned_type == score_type) |> 
  
  # Group by the assigned cell type to calculate our stats
  group_by(assigned_type) |> 
  
  # Calculate how many total bins, and how many have a score of exactly 0
  summarise(
    total_bins = n(),
    bins_with_zero_score = sum(score_value == 0, na.rm = TRUE),
    .groups = "drop"
  ) |> 
  
  # Add the percentage for context
  mutate(percentage_zero = (bins_with_zero_score / total_bins) * 100) |> 
  
  # Sort so the worst offenders are at the top
  arrange(desc(bins_with_zero_score))

# View the result
print(zero_score_summary)
```

Looks like is a unique problem of the Endothelial and MuSC (which are manually annotated)
DEVO RIMETTERE I MUSC??
```{r}
library(dplyr)

coords <- spatialCoords(spe) # get the coords
colnames(coords) <- paste0("spatial_", colnames(coords)) # make names unique

cd_df <- bind_cols(cd_df, as.data.frame(coords)) # prefer bind_cols over cbind
# cd_df <- cbind(cd_df, spatialCoords(spe))
valid_location_ids <- cd_df |> 
  select(location_id, cell_type, all_of(score_cols)) |> 
  pivot_longer(
    cols = all_of(score_cols), 
    names_to = "score_type", 
    values_to = "score_value"
  ) |> 
  # Keep only rows where the assignment matches the score column, AND the score is > 0
  filter(cell_type == score_type, score_value > 0) |> 
  pull(location_id) # This pulls just the location_ids into a simple vector

# 3. Filter your main dataframe using %in%
cd_df_clean <- cd_df |> 
  filter(location_id %in% valid_location_ids)

```


```{r}
cell_proportions <- cd_df_clean |> 
  count(cell_type) |> 
  mutate(
    proportion = n / sum(n),
    percentage = proportion * 100
  ) |> 
  # Sort so the most abundant cell type is at the top of the plot
  arrange(desc(percentage))

# plot
p_stacked <- cd_df_clean |> 
  ggplot(aes(x = "Filtered Tissue", fill = cell_type)) +
  
  # Removed color = "white" so the bars touch seamlessly
  geom_bar(position = "fill", width = 0.4) +
  
  # Applies your custom color palette
  scale_fill_manual(values = color_vector) +
  
  theme_minimal() +
  labs(
    title = "Composition of Confident Cell Types",
    subtitle = paste0("Total valid bins: ", nrow(cd_df_clean)),
    x = NULL, 
    y = "Proportion (0 to 1)",
    fill = "Cell Type"
  ) +
  theme(
    axis.text.x = element_text(size = 14, face = "bold"), 
    panel.grid.major.x = element_blank(), 
    panel.grid.minor.x = element_blank()
  )

print(p_stacked)
```

```{r}
cd_df_analyzed <- cd_df |> 
  rowwise() |> 
  mutate(
    # What was the score of the cell_type it was originally assigned?
    original_score = c_across(all_of(score_cols))[match(cell_type, score_cols)],
    
    # What is the highest score across ALL cell types for this bin?
    max_score = max(c_across(all_of(score_cols))),
    
    # WHICH cell type has that highest score?
    best_type = list(score_cols[which.max(c_across(all_of(score_cols)))])
  ) |> 
  ungroup()

# 3. Print the Diagnostic Summary!
cat("\n--- WHAT IS INSIDE THE SCORE=0 BINS? ---\n")
cd_df_analyzed |> 
  filter(original_score == 0) |> 
  mutate(
    # Group them into "Completely Empty" vs "Mislabeled"
    status = if_else(max_score == 0, 
                     "All 15 scores are 0", 
                     paste("Actually belongs to:", best_type))
  ) |> 
  count(assigned_as = cell_type, status, name = "number_of_bins") |> 
  arrange(desc(number_of_bins)) |> 
  print(n = 50) # Prints the top 50 rows so you can see where they went
```

#Funzione che assegna i nuovi tipi cellulari
```{r}
# 2. Define the function
fix_spe_celltypes <- function(spe, score_columns, cell_type_col = "cell_type") {
  
  # Extract colData as a dataframe
  cd <- as.data.frame(colData(spe))
  
  # Convert just the score columns into a numeric matrix (for extreme speed)
  score_mat <- as.matrix(cd[, score_columns])
  
  # A. Find the MAXIMUM score for each bin
  max_score <- apply(score_mat, 1, max)
  
  # B. Find WHICH cell type has that maximum score using max.col()
  best_type <- score_columns[max.col(score_mat, ties.method = "first")]
  
  # C. Find the score of the CURRENTLY assigned cell type
  # We find the column index of the assigned cell type, then extract the exact score
  orig_col_idx <- match(as.character(cd[[cell_type_col]]), score_columns)
  orig_score <- score_mat[cbind(1:nrow(score_mat), orig_col_idx)]
  
  # ---------------------------------------------------------
  # Apply your logic to create new_cell_type
  # ---------------------------------------------------------
  new_cell_type <- as.character(cd[[cell_type_col]])
  
  # Logic 1: If original score is 0, but another type is > 0 -> Reassign to best_type
  reassign_mask <- which(orig_score == 0 & max_score > 0)
  new_cell_type[reassign_mask] <- best_type[reassign_mask]
  
  # Logic 2: If EVERY score is 0 -> Assign NA
  na_mask <- which(max_score == 0)
  new_cell_type[na_mask] <- NA
  
  # Assign the new column back directly into the SPE object's colData
  spe$new_cell_type <- new_cell_type
  
  return(spe)
}

```

```{r}
# zeroscore_fun <- function(spe){
#   cd_df <- colData(spe) |> as_tibble()
#   
#   # zero score summary info
#   zero_score_summary <- cd_df |> 
#     select(location_id, assigned_type = cell_type, all_of(score_cols)) |> 
#     pivot_longer(
#       cols = all_of(score_cols), 
#       names_to = "score_type", 
#       values_to = "score_value"
#     ) |> 
#     filter(assigned_type == score_type) |> 
#     group_by(assigned_type) |> 
#     summarise(
#       total_bins = n(),
#       bins_with_zero_score = sum(score_value == 0, na.rm = TRUE),
#       .groups = "drop"
#     ) |> 
#     mutate(percentage_zero = (bins_with_zero_score / total_bins) * 100) |> 
#     arrange(desc(bins_with_zero_score))
#   
#   # Only return what is defined!
#   return(list(zeroScoreSummary = zero_score_summary))
# }
# zeroscore_list <- map(spe_list, zeroscore_fun)
# walk(zeroscore_list, \(el) {print(el$percentage_zero[1])})

zeroscore_fun <- function(subset){
  cd_df <- colData(subset) |> as_tibble()
  
  # zero score summary info
  zero_score_summary <- cd_df |> 
    select(barcode, assigned_type = cell_type, all_of(score_cols)) |> 
    pivot_longer(
      cols = all_of(score_cols), 
      names_to = "score_type", 
      values_to = "score_value"
    ) |> 
    filter(assigned_type == score_type) |> 
    group_by(assigned_type) |> 
    summarise(
      total_bins = n(),
      bins_with_zero_score = sum(score_value == 0, na.rm = TRUE),
      .groups = "drop"
    ) |> 
    mutate(percentage_zero = (bins_with_zero_score / total_bins) * 100) |> 
    arrange(desc(bins_with_zero_score))
  
  # Only return what is defined!
  return(list(zeroScoreSummary = zero_score_summary))
}
zeroscore_list <- map(subset_list, zeroscore_fun)
walk(zeroscore_list, \(el) {print(el$percentage_zero[1])})
```

# Analisi per tutta la lista dei campioni
```{r}
#spe_list <- map(spe_list, \(spe) {spe <- fix_spe_celltypes(spe, score_cols)})
subset_list <- map(subset_list, \(spe) {spe <- fix_spe_celltypes(spe, score_cols)})
map(subset_list, \(x){table(x$new_cell_type)})

#saveRDS(subset_list, file = "subset_list_updated.rds")

```

#proportions
```{r}
prop_df_ordered <- imap_dfr(subset_list_updated, function(spe, name) {
  tab <- base::table(colData(spe)[["new_cell_type"]])
  prop <- prop.table(tab)
  tissue_type <- unique(spe$tissue_type)
  data.table(
    sample_id = name,
    cell_type = names(prop),
    proportion = as.numeric(prop),
    tissue_type = as.character(tissue_type)
  )
}) |> 
  # lower case otherwise S is before f and m
  arrange(stringr::str_to_lower(tissue_type), sample_id) |>
  mutate(sample_id = fct_inorder(sample_id))

# vertical bars
ggplot(prop_df_ordered, aes(x = sample_id, y = proportion, fill = cell_type)) +
  geom_col() +                             # geom_col uses the y value directly (identity)
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = color_vector) +# use your named color_vector; otherwise use scale_fill_brewer()
  labs(x = "Sample", y = "Proportion", fill = "Cell type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

ggsave("~/prop_samples_updated.png", width = 10, height = 6, dpi = 300)
```

#proportions by conditions
```{r}
prop_df_condition <- imap_dfr(subset_list_updated, function(spe, name) {
  data.table(
    tissue_type = as.character(spe$tissue_type),
    cell_type = colData(spe)[["new_cell_type"]]
  ) |> 
    filter(!is.na(cell_type))
}) |> 
  arrange(stringr::str_to_lower(tissue_type)) |> 
  dplyr::count(tissue_type, cell_type) |>
  group_by(tissue_type) |>
  mutate(proportion = n / sum(n)) |>
  ungroup()


ggplot(prop_df_condition, aes(x = tissue_type, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per condizione",
    x = "Condizione (tissue_type)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("~/prop_updated.png", width = 10, height = 6, dpi = 300)
```

#Boxplots 
```{r boxplot-chunk, eval = !params$partial_data}
# This chunk will not run if partial_data == FALSE
prop_per_sample <- imap_dfr(spe_list, function(spe, name) {
  cd <- as.data.table(SummarizedExperiment::colData(spe))
  cd |>
    filter(!is.na(.data[["new_cell_type"]])) |>
    dplyr::count(.data[["new_cell_type"]], name = "count") |>
    mutate(proportion  = count / sum(count),
           sample      = name,
           tissue_type = paste(unique(as.character(cd$tissue_type)), collapse = ", ")) |> 
    dplyr::rename(cell_type = all_of("new_cell_type"))
})

ggplot(prop_per_sample, aes(x = tissue_type, y = proportion, fill = tissue_type)) +
  geom_boxplot(outlier.shape = 21) +
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1b9e77", "c26" = "#d95f02")) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribuzione proporzioni cell_type per tissue_type",
       x = "Tissue type",
       y = "Proporzione")

ggsave("~/repositories/cachetic_visiumHD/figures/output_R/emma_RCTD_by_me/boxplot_celltype_prop_by_cond.png", width = 10, height = 6, dpi = 300)
```

#Proiezioni sul tessuto
```{r}
subset_list_updated <- readRDS("~/subset_list_updated.rds")
# subset_list_updated <- lapply(subset_list_updated, function(spe) {
#    keep <- !is.na(spe$new_cell_type) & colSums(counts(spe)) > 0
#    spe[,keep]
# })

#saveRDS(subset_list_updated, file = "subset_list_updated.rds")

for (spe_name in names(subset_list_updated)) {
  spe <- subset_list_updated[[spe_name]]
  # ws <- colData(spe)[,seq(8,22,1)]
  # p <- lapply(names(ws), function(nm) {
  # plotCoords(spe, annotate = nm, point_size = 0.1) +
  #   scale_color_gradientn(
  #     colors = rev(hcl.colors(9, "Rocket")),
  #     limits = c(0, 1)  # fissa scala da 0 a 1
  #   ) +
  #   ggtitle(nm)
  # })

  # final_plot <- wrap_plots(p, nrow = 3, ncol = 5, guides = "collect") &
  # theme(
  #   legend.key.width = unit(0.5, "lines"),
  #   legend.key.height = unit(1, "lines"),
  #   strip.text = element_text(size = 5)
  # ) &
  # scale_color_gradientn(colors = rev(hcl.colors(9, "Rocket")), limits = c(0, 1))
  # 
  # ggsave(paste0(spe_name, "_types.png"), final_plot, width = 10, height = 8, dpi = 300)
  
your_color_vector <- c(
    "Endothelial" = "#0072B2",       # blu vivo
    "FAPs" = "#009E73",              # verde vivo
    "Immune_Cells" = "#D55E00",      # arancione-rosso
    "MuSC" = "#E69F00",              # arancione
    "Myonuclei_IIb" = "#CC79A7",     # magenta
    "Myonuclei_IIx" = "#56B4E9",     # azzurro chiaro
    "Myonuclei_IIx_IIa" = "#F0E442", # giallo
    "Myonuclei_IIx_IIb" = "#9B59B6", # viola chiaro
    "Myonuclei_MTJ" = "#0099B4",     # turchese
    "Myonuclei_NMJ" = "#DDAA33",     # senape
    "Myonuclei_Trim63" = "#CC3311",  # rosso scuro
    "Nervous_System" = "#44AA99",    # verde acqua
    "Pericyte" = "#AA4499",          # viola/rosa
    "Smooth_Muscular" = "#332288",   # blu-viola scuro
    "Tenocyte" = "#A3E635"           # verde lime
)

stat1 <- plotCoords(spe, annotate = "new_cell_type", point_size = 0.5) +
    scale_color_manual(values = your_color_vector) +
    theme(
      legend.key.width = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    )

ggsave(paste0(spe_name, "_dec.png"), stat1, width = 10, height = 8, dpi = 300)
}
  
```

#FOCUS SULLE IMMUNE CELLS
```{r}
#subset_list_updated <- readRDS("~/subset_list_updated.rds")

 default_color <- "lightgrey"
  your_color_vector <- setNames(
    rep(default_color, 2),
    c("Immune","other")
  )
  your_color_vector["Immune"]  <- "darkorange"

for (spe_name in names(subset_list_updated)) {
  spe <- subset_list_updated[[spe_name]]
  # ws <- colData(spe)[,seq(8,22,1)]
  # p <- lapply(names(ws), function(nm) {
  # plotCoords(spe, annotate = nm, point_size = 0.1) +
  #   scale_color_gradientn(
  #     colors = rev(hcl.colors(9, "Rocket")),
  #     limits = c(0, 1)  # fissa scala da 0 a 1
  #   ) +
  #   ggtitle(nm)
  # })

  # final_plot <- wrap_plots(p, nrow = 3, ncol = 5, guides = "collect") &
  # theme(
  #   legend.key.width = unit(0.5, "lines"),
  #   legend.key.height = unit(1, "lines"),
  #   strip.text = element_text(size = 5)
  # ) &
  # scale_color_gradientn(colors = rev(hcl.colors(9, "Rocket")), limits = c(0, 1))
  # 
  # ggsave(paste0(spe_name, "_types.png"), final_plot, width = 10, height = 8, dpi = 300)
  spe$imm_cell <- factor(ifelse(spe$new_cell_type == "Immune_Cells", 1, 0),
                       levels = c(0, 1),
                       labels = c("other", "Immune"))
  stat1 <- plotCoords(spe, annotate = "imm_cell", point_size = 0.9) +
    scale_color_manual(values = your_color_vector, na.value = default_color) +
    theme(
      legend.key.width = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    )
  ggsave(paste0(spe_name, "_imm_cells.png"), stat1, width = 10, height = 8, dpi = 300)
}

```

#Table to see proportions 
```{r}

#subset_list_updated <- readRDS("~/subset_list_updated.rds")

#by sample
celltype_table <- lapply(names(subset_list_updated), function(sample_name) {
  spe <- subset_list_updated[[sample_name]]
  df <- as.data.frame(colData(spe))
  df <- df %>%
    group_by(new_cell_type) %>%
    summarise(count = n()) %>%
    mutate(proportion = count / sum(count),
           sample = sample_name)
  df
}) %>%
  bind_rows()

celltype_matrix <- celltype_table %>%
  select(sample, new_cell_type, proportion) %>%
  pivot_wider(names_from = new_cell_type, values_from = proportion, values_fill = 0)

celltype_matrix
write_xlsx(celltype_matrix, "rctd16_prop_per_sample.xlsx")

#by condition
ct_long <- lapply(subset_list_updated, function(spe) {
  data.frame(
    tissue_type = unique(spe$tissue_type), 
    cell_type   = spe$new_cell_type
  )
}) %>%
  bind_rows()

counts <- ct_long %>%
  count(tissue_type, cell_type, name = "n")

props <- counts %>%
  group_by(tissue_type) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

props_wide <- props %>%
  select(tissue_type, cell_type, prop) %>%
  pivot_wider(names_from = cell_type, values_from = prop, values_fill = 0)

write_xlsx(props_wide, path = "rctd16_prop_cond.xlsx")

```

