---
title: "gfp_SpaceMosaic"
output: html_document
date: "2026-05-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The core idea is to define "patches" of cells that share a similar spatial context or exposure, perform DE analysis within these patches, and then aggregate the results using meta-analysis techniques.
This approach allows us to:
1.  Capture local spatial gradients that might be missed in global analyses.
2.  Account for spatial autocorrelation by modeling within contiguous units.
3.  Identify consistent biological responses across different regions of a tissue.

Goals
- Step 1: Create Patches: Build a flexible workflow to define spatial units using different inputs (e.g., cell neighborhoods, exposure hotspots).
- Step 2: Differential Expression: Perform patch-wise DE while considering spatial structure.
- Step 3: Meta-analysis: Summarize patch-level results to identify higher-level biological patterns and consistent gene responses.

```{r data-prep}
library(SpatialExperiment)
library(InSituCor)
library(ggplot2)
library(dplyr)
library(tidyr)
library(Matrix)
library(pheatmap)
library(ggrepel)

source("patchDE.R") #file in /per_git

```

```{r}
# Ensure results directory exists
if (!dir.exists("../results_spacemosaic/plots")) dir.create("../results_spacemosaic/plots", recursive = TRUE)

# Load the dataset
bin16_list_ann <- readRDS("~/bin16_list_ann.rds")

# Use one sample
spe_sub <- bin16_list_ann$c26murf1_b2

# Get spatial coordinates (convert to mm for patch definition)
xy <- as.matrix(spatialCoords(spe_sub))
xy_mm <- xy * 0.000316 # 1 pixel ~ 0,316 um
rownames(xy_mm) <- colnames(spe_sub)

```

Our biological question is: **How do epithelial cells behave in proximity of immune cells?** 
We define the "exposure" as the number of immune cells in the local neighborhood of each epithelial cell.
#maaa questa è un'altra cosa
#cioé qua la domanda è: come definisco i patches? in base a cosa?






