---
title: "Nuclei annotation with Rctd"
author: "Emma Menna"
output: html_document
date: "2026-03-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Libraries
```{r, include=FALSE}
library(arrow)
library(Banksy)
library(BiocParallel)
library(DropletUtils)
library(dplyr)
library(edgeR)
library(ggplot2)
library(graphics)
library(igraph)
library(jsonlite)
library(Matrix)
library(OSTA.data)
library(patchwork)
library(pheatmap)
library(S4Vectors)
library(scDblFinder)
library(scater)
library(scran)
library(scuttle)
library(SingleCellExperiment)
library(SpatialExperiment)
library(SummarizedExperiment)
library(tidyr)
library(VisiumIO)
library(spacexr)
library(SpotSweeper)
library(ggspavis)
library(writexl)
library(lme4)
library(purrr)
```

#Objects
```{r}
sce_ref <- readRDS("~/multiome_sce.rds") #reference dataset
nuclei_list_raw <- readRDS("~/nuclei_list_raw.rds")

color_vector <- c(
    "Endothelial" = "#0072B2",       # blu vivo
    "FAPs" = "#009E73",              # verde vivo
    "Immune_Cells" = "#D55E00",      # arancione-rosso
    "MuSC" = "#E69F00",              # arancione
    "Myonuclei_IIb" = "#CC79A7",     # magenta
    "Myonuclei_IIx" = "#56B4E9",     # azzurro chiaro
    "Myonuclei_IIx_IIa" = "#F0E442", # giallo
    "Myonuclei_IIx_IIb" = "#9B59B6", # viola chiaro
    "Myonuclei_MTJ" = "#0099B4",     # turchese
    "Myonuclei_NMJ" = "#DDAA33",     # senape
    "Myonuclei_Trim63" = "#CC3311",  # rosso scuro
    "Nervous_System" = "#44AA99",    # verde acqua
    "Pericyte" = "#AA4499",          # viola/rosa
    "Smooth_Muscular" = "#332288",   # blu-viola scuro
    "Tenocyte" = "#A3E635"           # verde lime
)

```

#Exploratory analysis on nuclei
```{r}
estrai_info <- lapply(nuclei_list, function(spe){
    n_spot_iniziale <- ncol(spe)
    n_spot_filtrati <- sum(spe$to_discard == "FALSE")
    total_umi <- sum(colSums(counts(spe)[, spe$to_discard == "FALSE"]))
    n_geni <- sum(rowSums(counts(spe)[, spe$to_discard == "FALSE"]) != 0)
    data.frame(
      Condition = unique(spe$sampletype),
      Block = unique(spe$blocco),
      Spot_number = n_spot_iniziale,
      Not_inflamed_spots = n_spot_filtrati,
      Total_UMI = total_umi,
      Detected_genes = n_geni
    )
})

tabella_finale <- do.call(rbind, estrai_info)
#write_xlsx(tabella_finale, path = "expl_analysis_nuclei.xlsx")

```

#Deconvolution function
Using "doublets" mode because these are not spots.
I set UMI_min very low considering the low counts in nuclei
and having assessed the difficulty of this algorithm in
attributing a significant annotation.
```{r}
run_rctd_doublet <- function(spe, sce_ref, out_prefix) {
  set.seed(9237)
  
  # data preparation
  rctd_data <- createRctd(
    spatial_experiment = spe,
    reference_experiment = sce_ref,
    cell_type_col = "subclusters",
    UMI_min = 10,
    UMI_min_sigma = 50,
  )
  
  # deconvolution in doublet mode
  results <- runRctd(rctd_data, rctd_mode = "doublet")

  pixels <- colnames(assay(results))
  
  first_type_map  <- setNames(colData(results)$first_type, pixels)
  second_type_map <- setNames(colData(results)$second_type, pixels)
  class_map       <- setNames(colData(results)$spot_class, pixels)
  
  spe$cell_type   <- first_type_map[as.character(spe$cell_id)]
  spe$second_type <- second_type_map[as.character(spe$cell_id)]
  spe$spot_class  <- class_map[as.character(spe$cell_id)]
  
  print(table(spe$spot_class, useNA = "ifany"))
  
  assign("nuclei_list", replace(nuclei_list, out_prefix, spe), envir = .GlobalEnv)
  
  invisible(spe)
}
```

#Rctd on all samples
```{r}
#changing assays name if needed
for (i in seq_along(nuclei_list)) {
  assayNames(nuclei_list[[i]])[assayNames(nuclei_list[[i]]) == "X"] <- "counts"
}

for (name in nuclei_list) {
    run_rctd_doublet(nuclei_list[[name]], sce_ref, name)
}
```

#Fixing rctd results
```{r}
#1. removing cells with NA cell type
nuclei_list <- lapply(nuclei_list, function(spe) {
   keep <- !is.na(spe$cell_type) & colSums(counts(spe)) > 0
   spe[,keep]
})

#2. manual correction on MuSC annotation: defined as Pax7 expressed or not
nuclei_list <- lapply(nuclei_list, function(spe) {
  print(table(spe$cell_type == "MuSC"))
  pax7_pos <- counts(spe)["Pax7", ] > 0
  spe$cell_type[pax7_pos] <- "MuSC"
  print(table(spe$cell_type == "MuSC"))
  return(spe)
})

#saving results
saveRDS(subset_list, file = "nuclei_list_ann.rds")
```

#projections on tissue with only significantly annotated nuclei
```{r}
for (spe_name in names(nuclei_list)) {
  spe <- nuclei_list[[spe_name]]
  spe <- spe[,spe$spot_class != "reject" & spe$to_discard == "FALSE"]
  print(dim(spe))
  spe$in_tissue <- rep(TRUE,dim(spe)[2])
  stat1 <- plotCoords(spe, annotate = "cell_type", point_size = 0.9,
                      x_coord = "y_coord", y_coord = "x_coord") +
    scale_color_manual(values = color_vector) +
    theme(
      legend.key.width = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    )

  ggsave(paste0(spe_name, "_dec.png"), stat1, width = 10, height = 8, dpi = 300)
}
  
```

#Barplot to show cell type proportions for each sample
```{r}

prop_df <- lapply(names(nuclei_list), function(name) {
  spe <- nuclei_list[[name]]
  spe <- spe[,spe$spot_class != "reject" & spe$to_discard == "FALSE"] #FACCIO QUA
  tab <- table(spe$cell_type)
  prop <- prop.table(tab)
  data.frame(
    sample = name,
    cell_type = names(prop),
    proportion = as.numeric(prop)
  )
}) |> bind_rows()

ggplot(prop_df, aes(x = sample, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per campione",
    x = "Campione (subset_list)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("cell_type_proportions_nuclei.png", width = 10, height = 6, dpi = 300)

```

#Barplot to show cell types proportions for each condition
```{r}
prop_df_condition <- lapply(nuclei_list, function(spe) {
  spe <- spe[, spe$spot_class != "reject" & !spe$to_discard]
  data.frame(
    tissue_type = unique(spe$sampletype),
    cell_type = spe$cell_type
  )
}) |> 
  bind_rows() |> 
  dplyr::filter(!is.na(cell_type))

prop_df_condition <- prop_df_condition %>% dplyr::count(tissue_type, cell_type) %>% dplyr::group_by(tissue_type) %>% dplyr::mutate(proportion = n / sum(n)) %>% ungroup()

ggplot(prop_df_condition, aes(x = tissue_type, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per condizione",
    x = "Condizione (tissue_type)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("cell_type_proportions_by_condition_nuclei.png", width = 10, height = 6, dpi = 300)

```

#Table to see proportions 
```{r}
#by sample
celltype_table <- lapply(names(nuclei_list), function(sample_name) {
  spe <- nuclei_list[[sample_name]]
  spe <- spe[, spe$spot_class != "reject" & !spe$to_discard]
  df <- as.data.frame(colData(spe))
  df <- df %>%
    group_by(cell_type) %>%
    summarise(count = n()) %>%
    mutate(proportion = count / sum(count),
           sample = sample_name)
  df
}) %>%
  bind_rows()

celltype_matrix <- celltype_table %>%
  select(sample, cell_type, proportion) %>%
  pivot_wider(names_from = cell_type, values_from = proportion, values_fill = 0)

write_xlsx(celltype_matrix, "celltype_prop_samples_nuclei.xlsx")

#by condition
ct_long <- lapply(nuclei_list, function(spe) {
  spe <- spe[, spe$spot_class != "reject" & !spe$to_discard]
  data.frame(
    tissue_type = unique(spe$sampletype),
    cell_type = spe$cell_type
  )
}) |> 
  bind_rows() |> 
  dplyr::filter(!is.na(cell_type))

counts <- ct_long %>%
  dplyr::count(tissue_type, cell_type, name = "n")

props <- counts %>%
  dplyr::group_by(tissue_type) %>%
  dplyr::mutate(prop = n / sum(n)) %>%
  dplyr::ungroup()

props_wide <- props %>%
  select(tissue_type, cell_type, prop) %>%
  pivot_wider(names_from = cell_type, values_from = prop, values_fill = 0)

write_xlsx(props_wide, path = "celltypes_conditions_nuclei.xlsx")

```

#boxplot of cell type proportions stratified by condition
```{r}
prop_per_sample <- purrr::imap_dfr(nuclei_list, function(spe, name) {
  spe <- spe[, spe$spot_class != "reject" & !spe$to_discard]
  cd <- as.data.frame(SummarizedExperiment::colData(spe))
  cd |>
    dplyr::filter(!is.na(cell_type)) |>
    dplyr::count(cell_type, name = "count") |>
    dplyr::mutate(
      proportion  = count / sum(count),
      sample      = name,
      tissue_type = paste(unique(as.character(cd$sampletype)), collapse = ", ")
    )
})
prop_per_sample$tissue_type <- factor(
  prop_per_sample$tissue_type,
  levels = c("sham", "c26", "c26STAT3", "c26foxO", "c26murf1", "c26SMAD23")
)
ggplot(prop_per_sample, aes(x = tissue_type, y = proportion, fill = tissue_type)) +
  geom_boxplot(outlier.shape = 21) +
  facet_wrap(~ cell_type, scales = "free_y") +
  scale_fill_manual(values = c("sham" = "#1B9E77", "c26" = "#D95F02")) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Distribuzione proporzioni cell_type per tissue_type",
       x = "Tissue type",
       y = "Proporzione")
ggsave("boxplot_cell_types.png", width = 10, height = 6, dpi = 300)

```

#ADDITIONAL ANALYSIS ---------------------------------------------------------------------------

#I wonder what cell types were assigned to the inflamed areas
```{r}
lapply(nuclei_list, function(spe){
  table(spe$cell_type[spe$to_discard=="TRUE"])
})

#plotting them
prop_df <- lapply(names(nuclei_list), function(name) {
  spe <- nuclei_list[[name]]
  spe <- spe[,spe$spot_class != "reject" & spe$to_discard == "TRUE"]
  if (ncol(spe) == 0) return(NULL)
  tab <- table(spe$cell_type)
  prop <- prop.table(tab)
  data.frame(
    sample = name,
    cell_type = names(prop),
    proportion = as.numeric(prop)
  )
}) |> bind_rows()

ggplot(prop_df, aes(x = sample, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per campione",
    x = "Campione (subset_list)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

ggsave("cell_type_proportions_nuclei.png", width = 10, height = 6, dpi = 300)

```

#Now I want to see which cell types the rejected (not significant) annotations would have been
```{r}

lapply(nuclei_list, function(spe){
  spe_filtered <- spe[,spe$to_discard=="FALSE"]
  #table(spe_filtered$spot_class!="reject")
  round(prop.table(table(spe_filtered$spot_class=="reject")),3)
})

lapply(nuclei_list, function(spe){
  table(spe$cell_type[spe$to_discard=="FALSE" & spe$spot_class=="reject"])
})

#we could look at how proportions change including these
prop_df <- lapply(names(nuclei_list), function(name) {
  spe <- nuclei_list[[name]]
  spe <- spe[,spe$to_discard == "FALSE"]
  if (ncol(spe) == 0) return(NULL)
  tab <- table(spe$cell_type)
  prop <- prop.table(tab)
  data.frame(
    sample = name,
    cell_type = names(prop),
    proportion = as.numeric(prop)
  )
}) |> bind_rows()

ggplot(prop_df, aes(x = sample, y = proportion, fill = cell_type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = color_vector) +
  theme_minimal() +
  labs(
    title = "Proporzioni dei tipi cellulari per campione",
    x = "Campione (subset_list)",
    y = "Proporzione",
    fill = "Tipo Cellulare"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

prop_df <- lapply(names(nuclei_list), function(name) {
  spe <- nuclei_list[[name]]
  spe <- spe[,spe$to_discard == "FALSE" & spe$spot_class =="reject"]
  if (ncol(spe) == 0) return(NULL)
  tab <- table(spe$cell_type)
  prop <- prop.table(tab)
  data.frame(
    sample = name,
    cell_type = names(prop),
    proportion = as.numeric(prop)
  )
}) |> bind_rows()


```

#Let's take a look at the second cell type as well
```{r}
lapply(nuclei_list, function(spe){
  table(spe$second_type[spe$to_discard=="FALSE" & spe$spot_class!="reject"])
})

#second cell type for each first one
lapply(nuclei_list, function(spe){
  spe_clean <- spe[,spe$to_discard=="FALSE" & spe$spot_class!="reject"]
  table(spe_clean$cell_type, spe_clean$second_type)
})

#plots
purrr::imap(nuclei_list, function(spe, sample_name) {
  spe <- spe[, !spe$to_discard & spe$spot_class != "reject"]
  if (ncol(spe) == 0) return(NULL)
  cd <- as.data.frame(SummarizedExperiment::colData(spe))
  cd <- cd %>%
    filter(!is.na(cell_type), !is.na(second_type))
  if (nrow(cd) == 0) return(NULL)
  df <- cd %>%
    dplyr::count(cell_type, second_type) %>%
    dplyr::group_by(cell_type) %>%
    dplyr::mutate(prop = n / sum(n)) %>%
    dplyr::ungroup()

  p <- ggplot(df, aes(x = "", y = prop, fill = second_type)) +
    geom_col(width = 1) +
    coord_polar(theta = "y") +
    facet_wrap(~cell_type) +
    scale_fill_manual(values = color_vector) +
    labs(title = sample_name) +
    theme_void()
  
  ggsave(
    paste0(sample_name, "_pie_second_type.png"),
    p,
    width = 10,
    height = 8,
    dpi = 300
  )
})

```

#Average on all samples
```{r}
all_props <- purrr::imap_dfr(nuclei_list, function(spe, sample_name) {
  
  spe <- spe[, !spe$to_discard & spe$spot_class != "reject"]
  if (ncol(spe) == 0) return(NULL)
  
  cd <- as.data.frame(SummarizedExperiment::colData(spe)) %>%
    filter(!is.na(cell_type), !is.na(second_type))
  
  if (nrow(cd) == 0) return(NULL)
  
  cd %>%
    count(cell_type, second_type) %>%
    group_by(cell_type) %>%
    mutate(prop = n / sum(n)) %>%
    ungroup() %>%
    mutate(sample = sample_name)
})

#media(normalizzata)
mean_props <- all_props %>%
  group_by(cell_type, second_type) %>%
  summarise(mean_prop = mean(prop), .groups = "drop") %>%
  group_by(cell_type) %>%
  mutate(mean_prop = mean_prop / sum(mean_prop)) %>%
  ungroup()

p <- ggplot(mean_props, aes(x = "", y = mean_prop, fill = second_type)) +
  geom_col(width = 1) +
  coord_polar(theta = "y") +
  facet_wrap(~cell_type) +
  scale_fill_manual(values = color_vector) +
  labs(title = "Mean proportions across samples") +
  theme_void()

ggsave(
  "mean_pie_second_type.png",
  p,
  width = 10,
  height = 8,
  dpi = 300
)
```

#Why are so many nuclei not significant?
```{r}
qc_list <- lapply(nuclei_list, function(spe){
  rejected <- spe[, spe$spot_class == "reject"]
  annotated <- spe[, spe$spot_class != "reject"]
  
  data.frame(
    total_counts = c(colSums(counts(rejected)), colSums(counts(annotated))),
    class = c(rep("reject", ncol(rejected)),
              rep("annotated", ncol(annotated)))
  )
})

qc_df <- dplyr::bind_rows(qc_list)

ggplot(qc_df, aes(x = class, y = total_counts)) +
  geom_violin() +
  geom_boxplot(width = 0.1) +
  scale_y_log10()

#####################
qc_list <- lapply(nuclei_list, function(spe){
  
  df <- data.frame(
    total_counts = colSums(counts(spe)),
    cell_type = spe$cell_type,
    spot_class = spe$spot_class
  )
  
  # assegna etichetta finale
  df$class <- ifelse(df$spot_class == "reject", 
                     "reject", 
                     as.character(df$cell_type))
  
  df
})

qc_df <- dplyr::bind_rows(qc_list)

ggplot(qc_df, aes(x = class, y = total_counts, fill = class)) +
  geom_violin(scale = "width") +
  geom_boxplot(width = 0.1, outlier.size = 0.3) +
  scale_y_log10() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#A separate barplot for each sample
```{r}

prop_df <- lapply(names(nuclei_list), function(name) {
  spe <- nuclei_list[[name]]
  spe <- spe[,spe$spot_class != "reject" & spe$to_discard == "FALSE"] #FACCIO QUA
  tab <- table(spe$cell_type)
  prop <- prop.table(tab)
  data.frame(
    sample = name,
    cell_type = names(prop),
    proportion = as.numeric(prop)
  )
}) |> bind_rows()

for (s in unique(prop_df$sample)) {
  
  df_sub <- prop_df %>% filter(sample == s)
  
  p <- ggplot(df_sub, aes(x = cell_type, y = proportion, fill = cell_type)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = color_vector) +
    theme_minimal() +
    labs(
      title = paste("Proporzioni tipi cellulari -", s),
      x = "Tipo Cellulare",
      y = "Proporzione"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      legend.position = "none"
    )
  
  ggsave(paste0("cell_type_proportions_", s, ".png"),
         plot = p, width = 8, height = 6, dpi = 300)
}

```

#Projections on tissue with not significant nuclei too
```{r}

for (spe_name in names(nuclei_list)) {
  spe <- nuclei_list[[spe_name]]
  spe <- spe[,spe$to_discard == "FALSE"]
  print(dim(spe))
  spe$in_tissue <- rep(TRUE,dim(spe)[2])
  stat1 <- plotCoords(spe, annotate = "cell_type", point_size = 0.9,
                      x_coord = "y_coord", y_coord = "x_coord") + #coordinate invertite!
    scale_color_manual(values = color_vector) +
    theme(
      legend.key.width = unit(0.5, "lines"),
      legend.key.height = unit(1, "lines")
    )

  ggsave(paste0(spe_name, "_dec.png"), stat1, width = 10, height = 8, dpi = 300)
}
  
```

#Barplot for comparisons between 16bins, all nuclei, sig nuclei
```{r}
names(bin16_list_ann)
names(nuclei_list_ann)
sham_c26_16bin <- bin16_list_ann[names(bin16_list_ann) %in% c("sham_b1", "c26_b2", "sham_b3",
                                                                        "c26_b4", "c26_b6", "sham_b6")]

sham_c26_nuclei <- nuclei_list_ann[names(nuclei_list_ann) %in% c("blocco1_sham", "blocco2_c26",
                                                                 "blocco3_sham", "blocco4_c26",
                                                                 "blocco6_c26", "blocco6_sham")]
names(sham_c26_nuclei) <- c("sham_b1", "c26_b2", "sham_b3", "c26_b4", "sham_b6", "c26_b6")

plot_list <- lapply(names(sham_c26_nuclei), function(name) {
  
  nuclei <- sham_c26_nuclei[[name]]
  bin16  <- sham_c26_16bin[[name]]
  
  # ---- 1. 16bin ----
  tab1 <- table(colData(bin16)$new_cell_type)
  prop1 <- prop.table(tab1)
  
  df1 <- data.frame(
    source = "16bin",
    cell_type = names(prop1),
    proportion = as.numeric(prop1)
  )
  
  # ---- 2. nuclei clean ----
  nuclei_all <- nuclei[, nuclei$to_discard == FALSE]
  tab2 <- table(colData(nuclei_all)$cell_type)
  prop2 <- prop.table(tab2)
  
  df2 <- data.frame(
    source = "nuclei_all",
    cell_type = names(prop2),
    proportion = as.numeric(prop2)
  )
  
  # ---- 3. nuclei filtered ----
  nuclei_sig <- nuclei[, nuclei$to_discard == FALSE & nuclei$spot_class != "reject"]
  tab3 <- table(colData(nuclei_sig)$cell_type)
  prop3 <- prop.table(tab3)
  
  df3 <- data.frame(
    source = "nuclei_sig",
    cell_type = names(prop3),
    proportion = as.numeric(prop3)
  )
  
  df_plot <- rbind(df1, df2, df3)
  
  # ordine barre
  df_plot$source <- factor(df_plot$source,
                           levels = c("16bin", "nuclei_all", "nuclei_sig"))
  
  # PLOT (QUESTO è quello che volevi)
  ggplot(df_plot, aes(x = source, y = proportion, fill = cell_type)) +
    geom_bar(stat = "identity") +
    scale_fill_manual(values = color_vector) +
    theme_minimal() +
    labs(
      title = name,
      x = NULL,
      y = "Proporzione",
      fill = "Cell type"
    )
})

# esempio: primo campione
plot_list[[2]]
```



